\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\input{included_packages} % Required for inserting images

\title{NUR A Hand in 1}
\author{Matthew Kirwan \\
        S4557093}
\date{March 2026}

\begin{document}

\maketitle

\section*{Introduction}
This study examines the use of memory efficient methods with the aim of handling overflow and underflow in the Poisson distribution. Additionally, we implement some linear algebra methods to obtain the 19 degree polynomial from the Vandermonde matrix, and comparing the error against the number of iterations of the LU Decomposition and Nevilles algorithm.

\section*{Problems and Methods}
In this section, the steps taken to solve the given problems are explained. We begin by looking at the initialization of the script.

We import a number of modules for data/path handling, graphing, and timing purposes. Here we automatically grab the directory the script is in, and navigate around this path to access the directory structure surrounding it.

\begin{lstlisting}[language=Python]
import os
import time
import math
import numpy as np
import pandas as pd

from functions import interpolation
from functions import math_functions
from functions import linear_algebra as la

import matplotlib.pyplot as plt

# get the folder where the script is located
script_dir = os.path.dirname(__file__)
data_dir = os.path.join(script_dir, "../data")
plots_dir = os.path.join(script_dir, "../plots")

os.makedirs(plots_dir, exist_ok=True)  # just in case
\end{lstlisting}

In the section above we import 3 files which will be used in solving the problems proposed, 'interpolation', 'math\_functions', and 'linear\_algebra'. We will explore these further upon their usage.

\subsection*{Question 1: Poisson Distribution}
The problem proposed here is focused around handling over/underflow through the calculation of the Poisson distribution, defined in eq. \ref{eq:poisson}

\begin{equation}
    P_\lambda(K) = \frac{\lambda^ke^{-\lambda}}{k!}
    \label{eq:poisson}
\end{equation}

Here we are restricted to the use of 32 bit variables, so we utilise the numpy.float32 and numpy.int32 functions to ensure this restriction is met. In order to test the effectiveness of our algorithm, we are given a number of values which are defined in table \ref{tab:lam_and_k}. We are also expected to output to at least six significant figures.

\begin{table}[]
    \centering
    \begin{tabular}{|c|c|}
        \hline
        $\lambda$ & k \\
        \hline
        1 & 0 \\
        5 & 10 \\
        3 & 21 \\
        2.6 & 40 \\
        100 & 5 \\
        101 & 200 \\
        \hline
    \end{tabular}
    \caption{$\lambda$ and k values given}
    \label{tab:lam_and_k}
\end{table}

We begin by defining the test data in the question\_1\_values numpy array, ensuring the data is stored in numpy.float32 bits. We then then define an empty list for appending our results to later and begin looping through the test variables. In order to handle both the integer and float $\lambda$ values, we load them in as floats initially, and check if the float value is an integer, if it is we convert this $\lambda$ to an numpy.int32 variable, otherwise we do nothing. We then ensure that k is an numpy.int32 variable as the factorial can only ever handle integers.

We can now safely pass the values through to the Poisson distribution defined in the math\_functions, this returns a single scalar value which we then append to the results list as an array and then convert this to a pandas DataFrame for pretty printing.

\begin{lstlisting}[language=Python]
def question_1() -> pd.DataFrame:
    question_1_values = np.array([[1,0],[5,10],[3,21],[2.6,40],[100,5],[101,200]],dtype=np.float32)

    results = []
    for values in question_1_values:

        #defining how the data is stored in 32 bits
        lam_raw = np.float32(values[0])
        if lam_raw.is_integer():
            lam = np.int32(lam_raw)
        else:
            lam = lam_raw
        k = np.int32(values[1])

        result = math_functions.poisson_distribution(lam,k)
        results.append([lam,k,result])
    results = pd.DataFrame(results,columns=['lambda','k','Poisson result'])
    return results
\end{lstlisting}

When we pass the variables to the Poisson distribution function, we check if k is 0, in the case it is, we calculate the distribution as it is written in eq. \ref{eq:poisson}. This is truly just an artifact of early prototyping, as the other condition can handle the log(0) due to the fact that there is no division. In the other case, we apply a logarithm to the entire distribution, which results in the equation to be re-written as
eq. \ref{eq: log_poisson}. Once the values are obtained, they are returned as numpy.float32 variables.

\begin{equation}
    Log(P_{(\lambda,k)}) = kLog(\lambda) - \lambda - Log(k!)
    \label{eq: log_poisson}
\end{equation}

\newpage
\begin{lstlisting}[language=Python]
def poisson_distribution(lam:np.float32|np.int32,
                         k:np.int32
                         ) -> np.int32:

    if k == 0:
        p = ((lam ** k) * np.exp(-lam)) / (factorial(k))
        return np.float32(p)
    else:
        p = k * math.log(lam) - lam - log_factorial(k)
        return np.exp(np.float32(p))
\end{lstlisting}

We can encounter overflow easily if k is double digits, to handle this, we implement a logarithmic factorial function. Regardless, the factorial function is a recursive function that returns the product of the previous factorials once 1 is reached. These values are returned in numpy.float32 variables so that they can be returned as an exponent.

The log factorial function is not recursive but a for loop which obtains the summation of the Log of each value until the k integer is reached. As we initialize the sum with 0, and don't handle the log of 0, this function can easily handle the k = 0 case. As we require the log of the factorial, this value is returned as a numpy.float32 variable.

\begin{lstlisting}[language=Python]
#tried to use this initially but failed at k = 40 due to overflow
#tried return log(product) but realised that was a recursive log return so...
def factorial(x:np.int32
              ) -> np.float32:

    if x > 1:
        return np.float32(x * factorial(x-1))
    else:
        return np.float32(1)

#solved overflow issue
def log_factorial(x:np.int32
                  ) -> np.float32:

    fac_sum = 0.0
    for i in range(1, x + 1):
        fac_sum += math.log(i)

    return np.float32(fac_sum)
\end{lstlisting}

\subsection*{Question 2: Vandermonde Matrix}
In this section we are first challenged with obtaining the Vandermonde Matrix \textbf{V}, which is defined by eq. \ref{eq: vandermonde}. Where $x_i$ is 20 points defined in the vandermonde.txt file. Secondly we are asked to obtain the 19 degree polynomial for these 20 points through the use of LU decomposition and 19 degree polynomial interpolation, and timing the time required for each process.

\begin{equation}
    V_{i,j} = x_i^j
    \label{eq: vandermonde}
\end{equation}

I've removed some sections from this function for ease of reading (ie. plotting, saving) and the return declaration is incorrect here, as we return a tuple of multiple arrays. We start this section by importing the data from the vandermonde.txt file. Splitting the data and reorganising the data in to more usable formats, and initialise our storage.

We then calculate the vandermonde matrix with the math\_functions function get\_vandermonde\_matrix(x). Once we have this we can perform our LU decompositions over 1 iteration and 10 iterations with the functions LU\_n\_iter and can time this process with the perf\_counter, we opted out of the usage of timeit as we require the outputs of these functions.

We then perform a timing analysis on the nevilles algorithm implementation by parsing through the 1001 x points and interpolating their values. This function utilises the M variable which is the amount of points around the requested data point (found through bisection) that will be considered during the interpolation, setting M = 20, means all points are considered at all interpolations resulting in a 19 degree polynomial fit. Setting this value lower to 3/5/8 allows for a much smoother approximation as not all points are considered, only neighbouring points.

After this we interpolate the values for all 1001 points through linear/lagrange/Nevilles algorithm interpolation. These values are saved and later plotted. Subsequently, we re-interpolate values for the known x values and obtain the difference between the known value $y_i$ and the interpolated value y(x)

\begin{lstlisting}[language=Python]
def question_2() -> tuple[pd.DataFrame]:
    data = import_data()
    x,y = np.array(data['x']),np.array(data['y'])
    x_points_to_interpolate = np.linspace(x[0],x[-1],1001)
    M=20 #change to 3/5/8 for smooth function
    nevilles    = []
    lagrange    = []
    linear      = []
    x_points = []
    x_y = np.array([x, y])

    vandermonde = math_functions.get_vandermonde_matrix(x)

    start = time.perf_counter()
    c, ys = LU_1_iter(vandermonde, x, y)
    end = time.perf_counter()
    elapsed_1_iter = end - start

    start = time.perf_counter()
    c_1, yyy = LU_10_iter(vandermonde, x, y)
    end = time.perf_counter()
    elapsed_10_iter = end - start

    start = time.perf_counter()
    for point_index in range(0,len(x_points_to_interpolate)):

        point_request = x_points_to_interpolate[point_index]
        interpolated_values_nev     = interpolation.nevilles_algorithm(point_request,M,x_y)
    end = time.perf_counter()
    elapsed_nev = end - start

    #need to split this up for timing purposes
    for point_index in range(0,len(x_points_to_interpolate)):
        point_request = x_points_to_interpolate[point_index]
        interpolated_values_linear  = interpolation.linear_interpolation(point_request,x_y)
        interpolated_values_nev     = interpolation.nevilles_algorithm(point_request,M,x_y)
        interpolated_values_lag     = interpolation.lagrange_polynomial(point_request,19,x_y)
        linear.append(interpolated_values_linear[1])
        nevilles.append(interpolated_values_nev[1])
        lagrange.append(interpolated_values_lag[1])
        x_points.append(x_points_to_interpolate[point_index])

    y_check=[]
    for point_request in x:
        interpolated_values_nev = interpolation.nevilles_algorithm(point_request,M,x_y)
        y_check.append(interpolated_values_nev[1])
    diff_nev = abs(y_check - y)

    y_check = []
    c_1, y_check = LU_1_iter(vandermonde, x, y,20)
    diff_1st = abs(y_check - y)

    y_check = []
    c_1, y_check = LU_10_iter(vandermonde, x, y,20)
    diff_10th = abs(y_check - y)

\end{lstlisting}

We call multiple functions here, so we will examine what they do. The first is the import data function which is rather simple, we just import the data as a pandas DataFrame from a location on the repo.

\begin{lstlisting}[language=Python]
def import_data():

    data_path = os.path.join(data_dir, "Vandermonde.txt")
    data=pd.DataFrame(np.genfromtxt(data_path,comments='#',dtype=np.float64),columns=['x','y'])
    return data
\end{lstlisting}

We then get the Vandermonde Matrix, performs an implementation of eq. \ref{eq: vandermonde}, and was taken from my work last year.

\begin{lstlisting}[language=Python]
def get_vandermonde_matrix(x):
    # taken from last years work
    V_m = np.zeros((len(x),len(x)))
    for j in range(V_m.shape[1]):
        for i in range(V_m.shape[0]):
            V_m[i,j] = x[i]**j
    return V_m
\end{lstlisting}

We then implement the multiple iterations of the LU decomposition. Unfortunately I did not quite know how to perform this decomposition recursively or have time to ask for help, so the 10 iteration attempt is what I have tried to do to minimise error by getting the residual error in the y calculation, and adjusting the c vector with this. These functions call upon the linear algebra package and utilises the Crout's algorithm, forward and backward substitution, and the math function get y which is my old version of what is shown in LU 10 iters.

\begin{lstlisting}[language=Python]
def LU_1_iter(vandermonde,x,y,points = 1001):
    A = la.crouts_algorithm(vandermonde)
    y1 = y.copy()
    out = la.forward_sub(A, y1)
    c = la.backward_sub(A, out)
    y = math_functions.get_y(x,c,points)
    return c, y

def LU_10_iter(vandermonde,x,y,points = 1001):
    y1 = y.copy()
    A = la.crouts_algorithm(vandermonde)
    z = la.forward_sub(A, y1)
    c = la.backward_sub(A, z)

    for _ in range(100):
        residual = y1 - vandermonde @ c
        z = la.forward_sub(A, residual)
        delta = la.backward_sub(A, z)
        c += delta
    y = math_functions.get_y(x,c,points)
    return c, y
\end{lstlisting}

Crout's algorithm is implemented from my understanding of the lecture notes, I found that my first interpretation did not achieve good benchmark results for the 19 degree polynomial, but did obtain good benchmark results on small 3x3 matrices. So I tried a new attempt, in which I split up my j loop to handle the top and bottom triangular matrices separately, with the bottom loop handling the $\alpha$ matrix and the top loop handling the $\beta$ matrix. Admittedly, I should have overwrote A, however I was worried about requiring the information later. I tired to avoid a third loop by utilizing smart indexing which I believe has worked as intended but I will need to research this further and ask questions about the most optimal solution.

\newpage
\begin{lstlisting}[language=Python]
def crouts_algorithm(matrix_A):

    A = matrix_A.copy()
    B = np.zeros((len(A),len(A.T)))
    if len(A.T) != len(A):
        raise Warning('Matrix is not square')

    for i in range(len(A)):
        # for j in range(len(A.T)):
        #     if i == 0:
        #         B[i,j] = A[i,j]
        for j in range(i,A.shape[0]):
            B[j,i] = A[j,i]-np.sum(B[j,:i]*B[:i,i])
        for j in range(i+1,A.shape[0]):
            B[i,j] = (1/B[i,i])*(A[i,j]-np.sum(B[i,:i]*B[:i,j]))

    #B is the LU decomp
    return B

def crouts_algorithm_first(matrix_A):
    A = matrix_A.copy()
    B = np.zeros((len(A),len(A.T)))
    if len(A.T) != len(A):
        raise Warning('Matrix is not square')

    for i in range(len(A)):
        B[i, i] = 1
        for j in range(len(A.T)):
            if i == 0:
                B[i,j] = A[i,j]
            else:
                # checked the result of this against this link, so I think its right
                # www.emathhelp.net/calculators/linear-algebra/lu-decomposition-calculator
                if i <= j:
                    B[i,j] = A[i,j] - np.sum(B[i,:j]*B[:j,j])
                if i > j:
                    B[i,j] = (1/B[j,j])*(A[i,j]-np.sum(B[i,:j]*B[:j,j]))

    #B is the LU decomp
    return B
\end{lstlisting}

To utilise the LU decomp to find the solutions we also have to implement forward and backward substitution. These functions were improvements on my code from last year avoiding the unnecessary inverse in the backward substitution.

\begin{lstlisting}[language=Python]
def forward_sub(matrix_A,y):
    z = np.zeros(matrix_A.shape[0])
    for i in range(matrix_A.shape[0]):
        z[i] = (1/matrix_A[i,i])*(y[i]-matrix_A[i,:i]@z[:i])
    return z

def backward_sub(matrix_A,z):
    c = np.zeros(matrix_A.shape[0])
    for i in range(matrix_A.shape[0]-1,-1,-1):
        c[i] = (z[i] - (matrix_A[i,i+1:]@c[i+1:]))
    return c
\end{lstlisting}

My get y function is ripped from my work last year, I just tried to implement a smarter version and realised c is limited to 20, where as the V matrix would be 1000x1000. So I think this works for now as a way to step by step pass through the polynomial. Perhaps this is the source of error for the LU decomp and a smarter implementation would reduce this significantly.
\newpage

\begin{lstlisting}[language=Python]
def get_y(x,c,n):
    #is this wrong?
    #this seems fine...
    new_x = np.linspace(x[0],x[-1],n)
    new_y = np.zeros(len(new_x))

    for i in range(len(new_x)):
        total = 0
        for j in range(c.shape[0]):
            total+=c[j]*(new_x[i]**j)
        new_y[i]=total
    return new_y
\end{lstlisting}

From here we move onto the interpolation algorithms. I will start by introducing my bisection algorithm which is not simple or elegant but I tested to catch edge cases. It handles monotonic values, multiple edge samples, single edge samples, assumes the left case is dominant and handles poor inputs. I think this is fairly robust and works well with the interpolation schemes I have implemented.

\begin{lstlisting}[language=Python]
def bisection(point_request:float,
              data_input:np.ndarray,
              points_around:int|None = None #this is value left/right ie 2 would be n-2 n-1 n n+1 n+2
              ) -> tuple[np.array,bool]:

    data = data_input.copy()
    edge_missing = False
    try:
        left_index = 0
        right_index = len(data[0]) - 1
        left_edge = data[0,left_index]
        right_edge = data[0,right_index]
    except:

        left_index = 0
        right_index = len(data) - 1

        left_edge = data[left_index]
        right_edge = data[right_index]

    # assume point is within range
    assert left_edge <= point_request <= right_edge, ("point out of range")

    #assume left dominant
    middle_index = math.floor((right_index-left_index) / 2)
    #continue until edges found
    while right_index - left_index > 1:

        # assume left dominant, update edges
        try:
            if point_request <= data[0,middle_index]:
                right_index = middle_index
            elif point_request > data[0,middle_index]:
                left_index = middle_index
        except:
            if point_request <= data[middle_index]:
                right_index = middle_index
            elif point_request > data[middle_index]:
                left_index = middle_index
        middle_index = math.floor((left_index+right_index)/2)

        #check stopping condition
        if right_index == left_index+1:

            #check if point falls on one of the edges, if so both points the same
            try:
                if point_request == data[0,left_index]:
                    right_index = left_index
                elif point_request == data[0,right_index]:
                    left_index = right_index
            except:
                if point_request == data[left_index]:
                    right_index = left_index
                elif point_request == data[right_index]:
                    left_index = right_index

    #if no boundary condition return edges


    if points_around is None:
        return np.array([data[0][left_index:right_index+1],data[1][left_index:right_index+1]]), edge_missing

    #else return list with boundary indices included
    else:

        assert points_around > 0, ('Points around is negative, required to be stritcly postitive')
        left_bound = left_index - points_around
        right_bound = right_index + points_around
        if left_bound < 0:
            left_bound = 0
            edge_missing = True
        if right_bound > len(data[0])-1:
            right_bound = len(data[0])-1
            edge_missing = True

        return np.array([data[0][left_bound:right_bound+1],data[1][left_bound:right_bound+1]]), edge_missing

\end{lstlisting}

Once the bisection algorithm was implemented we could implement the following codes. Linear interpolation was straight forward, it just required weighing the y value from the left and right point against the proposed x point, drawing a straight line between points. If the points were monotonic, the y value for that point was returned.

\begin{lstlisting}[language=Python]
def linear_interpolation(point_request:float,
                         data = np.ndarray,
                         ) -> np.ndarray[float,float]:

    points, egde = bisection(point_request,data)
    if points.shape[1] == 1:
        return np.array([points[0,0],points[1,0]])
    else:
        right_point = points[:,0]
        left_point = points[:,1]
    fraction = (right_point[1] - left_point[1])/(right_point[0] - left_point[0])
    interpolated_value = fraction*(point_request - left_point[0]) + left_point[1]
    return np.array([point_request,interpolated_value])

\end{lstlisting}

We then implemented the Lagrange polynomial to check if our algorithms were benchmarking against each other. There was a assertion statement to ensure the degree of the polynomial did not pass 5, however we needed to check for a degree of 19, so this was silenced. I did a product of the equation given in the lectures and then summed the products to get the full polynomial.
\newpage

\begin{lstlisting}[language=Python]
def lagrange_polynomial(point_request:float,
                        degree_of_polynomial:int,
                        full_data:np.ndarray
                        ) -> np.ndarray[float,float]:
    # assert degree_of_polynomial <= 5, ('Degree of polynomial should be less than or equal to 5')
    assert degree_of_polynomial > 0, ('Degree of polynomial must be strictly postitive and greater than 0')

    data = full_data.copy()

    nearby_samples, edge_case = bisection(point_request,data,40)
    # if edge_case:
    #     return np.array([point_request, None])
    nearby_data = nearby_samples.copy()
    interpolated_value = 0

    for i in range(degree_of_polynomial+1):
        products = nearby_data[1,i]
        for j in range(degree_of_polynomial+1):
            if j == i:
                continue
            product = (point_request - nearby_data[0,j])/(nearby_data[0,i] - nearby_data[0,j])
            products *=product

        interpolated_value += products

    return np.array([point_request, interpolated_value])
\end{lstlisting}

We then made an implementation of Neville's algorithm to generate a polynomial of any degree allowed given the data available. I implemented a version originally that did not work due to the indicies and signs. I also had a closest pair function as stated in the slides but could not figure out where to implement it so this was removed.

\begin{lstlisting}[language=Python]
def nevilles_algorithm(point_request:float,
                       M_points_around:int,
                       data:np.ndarray
                       ) -> np.ndarray[float,float]:

    nearby_samples, edge_case = bisection(point_request,data,M_points_around)
    #save only ys
    p_i = nearby_samples[1,:].copy()

    # tried to follow the slides perfectly, used https://github.com/gisalgs/geom/blob/master/neville.py
    # to adjust my indices and sums
    for k in range(1,len(p_i)):
        for i in range(len(p_i)-k):
            j= i + k
            top_1 = (point_request - nearby_samples[0,j])*p_i[i]
            top_2 = (nearby_samples[0,i] - point_request)*p_i[i+1]
            bottom = 1/(nearby_samples[0,i]-nearby_samples[0,j])
            p_i[i] = (top_1+top_2)*bottom

    return np.array([point_request,p_i[0]])

    # heres what I had before not super different just wasn't summing correct
    # def nevilles_algorithm(point_request:float,
    #                        M_points_around:int,
    #                        data:np.ndarray
    #                        ) -> tuple[float,float]:
    #     nearby_samples, edge_case = bisection(point_request,data,M_points_around)
    #     print(nearby_samples)
    #     #save only ys
    #     p_i = nearby_samples[:,1].copy()
    #     #find the closest pair or single x value
    #     nearest_pair, _ = bisection(point_request,data)
    #     if len(nearest_pair[:,0]) > 1:
    #         closest = check_closest(nearest_pair,point_request)
    #         initial_solution = closest[1]
    #     else:
    #         initial_solution = nearest_pair[0,1]
    #
    #     for k in range(1,len(p_i)):
    #
    #         print(p_i)
    #         for i in range(0,len(p_i)-k):
    #             j= i + k
    #             top_1 = (nearby_samples[j,0]-point_request)*p_i[i]
    #             top_2 = (point_request-nearby_samples[i,0])*p_i[j]
    #             bottom = 1/(nearby_samples[j,0]-nearby_samples[i,0])
    #             p_i[i] = (top_1+top_2)*bottom
    #     print(p_i)
    #     print(point_request)
    #     print(nearby_samples)
    #     print(p_i)
\end{lstlisting}



\section*{Results}
\subsection*{Question 1: Poisson Distribution}

We find that the script returns a value for every test input without any hardcoding, and we believe this achieves a high accuracy at returning 7 significant figures. The results of this can be seen in table. \ref{}

\input{../data/poisson.csv}

\subsection*{Question 2: Vandermonde Matrix}

\newpage
\appendix
\section*{Appendix}

I made a new implementation of the Gauss-Jordan elimination algorithm which I could not find a purpose for here.
\begin{lstlisting}[language=Python]
def gauss_jordan_elimination(matrix_A:np.ndarray,
                             vector_b:np.ndarray
                             )->tuple[np.ndarray,np.ndarray]:

    # b is the solution
    #could include the identity for the inverse
    A = matrix_A.copy()
    b = vector_b.copy()

    if len(A.T) != len(A):
        raise Warning('Matrix is not square')
    for i in range(len(A.T)):
        #i is column index
        # A[:,i] is column
        largest_index = None
        pivot = 0

        for j in range(i,len(A)):

            #j is from column index to row index max
            # loop over rows from row i
            # find max in column i
            # so A[j,i]
            if abs(A[j,i]) > abs(pivot):
                largest_index = j
                pivot = A[j,i]

        if abs(pivot) > 0:

            #swap then normalise
            A[[i, largest_index]]   = A[[largest_index, i]]
            A[i,:]                  = A[i,:]/pivot

            b[[i,largest_index]]    = b[[largest_index,i]]
            b[i]                    = b[i]/pivot

            #reduce above and below diagonal to 0
            # need to make a copy first for the b vector

            above_terms = A[:i, i].copy()
            below_terms = A[i+1:, i].copy()
            A[:i, :]    -= above_terms[:, None]     * A[i, :]
            A[i+1:, :]  -= below_terms[:, None]     * A[i, :]

            b[:i]       -= above_terms              * b[i]
            b[i+1:]     -= below_terms              * b[i]

        else:
            raise Warning('Matrix is not singular')

    return b
\end{lstlisting}


\end{document}
